---
title: "Day 5"
---

## Goal

Fork a repo which already includes the data which we used yesterday. Connect to a project on your computer. Make a website with `quarto create project website .` Put a graphic in the `index.qmd` file. Finish with a URL in #class-work.

## Session 1

Show how to fork: https://github.com/bootcamp-2025-06/nyc-census-tracts-to-fork

After you fork, you can (and should) change the name.

## Session 2

Cold call: Walk me through the initial stages.

## Session 3

Cold call: Finish up the initial steps.

Did you change the .gitignore? Including the _files? Did you push to Github? 

Make sure that everyone in your group has done this.

## Session 4

Let's take a closer look at the data. (We should have done this yesterday.)

What are some commands to look at a new data set? How about print(), summary(), slice_sample(n = 5), tail()? Maybe skimr::skim?

Can we create a new county variable? 

What do the words in tract mean? How many counties are there in NYC? And do those county names make sense?

## Session 5

Cold call: What did you group do?

Make a plot. Don't make it the same as yesterday! Do something different. How about overlapping density plots? Maybe a box-plot by race? Maybe a plot which shows a bar plot with the standard deviation for each race? Maybe a histogram of med_income, faceted by race? 

```{r}
#| message: false
#| echo: false
#| label: setup

library(tidyverse)

x <- read_csv("data/nyc_tracts.csv", show_col_types = FALSE)
```

```{r}
x |> 
  mutate(race = factor(race, levels = c("Hispanic", "Black", 
                                         "Asian", "White"))) |> 
  drop_na() |>                                        
  ggplot(aes(x = med_income, fill = race)) +
    geom_density(alpha = 0.5, adjust = 2) +
    theme_minimal() +
    scale_x_continuous(labels = scales::dollar_format()) +
    theme(axis.text.y = element_blank()) +
    labs(title = "Median Income for NYC Census Tracts",
         subtitle = "The richest tracts are White",
         x = "Median Income",
         y = NULL,
         fill = "Race",
         caption = "The race of each tract is determined by the most numerous racial group\n present in that tract. Data from the American Community Survey 2022.")
```

## Session 6

Cold call: How are things going?

## Session 7

Publish the plot to GitHub Pages. Looks OK. But it is a little ugly to have the title and, maybe the author, edit the YAML to remove. Re-publish. Add link to class-work.

## Extra Questions

Create a new county variable. Give them the correct (famous) names. Try faceting the plots. Ask AI to explain. Ask AI to write some text.


