---
title: "Wisdom and Governors"
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(gt)
library(primer.data)

```

## Background Information

This dataset is drawn from an attempt to estimate the causal effect of winning an election on the subsequent lifespan of political candidates. The study includes data from candidates who participated in competitive elections, with the treatment defined as winning versus losing a narrowly decided race. By focusing on close contests, the analysis aims to approximate random assignment, helping to address confounding variables that might otherwise bias estimates. The dataset records each candidate’s age at election, party affiliation, gender, vote margin, and eventual age at death, allowing for regression-based modeling of the relationship between electoral outcomes and longevity. 

Commentary: http://www.stat.columbia.edu/~gelman/research/published/causal_paths_3.pdf
Response: https://erikgahner.dk/2020/a-response-to-andrew-gelman/

### Scenarios

1) Imagine you work for a life insurer, and want to forecast how long candidates for the US Senate might live based on their age, party, election result, and other variables. 

2) Imagine you are a researcher. You want to know if winning candidates live longer. 

### Data

For this assignment, we will use the `governors` dataset, which contains information on U.S. gubernatorial candidates and their election outcomes. This dataset comes from published research and is available in the `primer.data` package.

* Explore the data. What is the outcome variable? How is it distributed?

```{r}
governors
```

```{r}
summary(governors)
table(governors$party)
table(governors$region)
table(governors$sex)
```

```{r}
governors |> 
  ggplot(aes(x = death_age)) + 
    geom_histogram(bins = 30) +
      coord_cartesian(xlim = c(0, 110))
```


### Preceptor Table

What sorts of questions are we trying to answer?

In order to build the Preceptor Table, we need to know: Is the model causal or predictive? What are units? What are the outcomes? What are the covariates? What are the treatments? Describe the mathematical structure of the model.

* Causal or predictive?

Scenario 1: Predictive — We are forecasting the expected lifespan of current political candidates using their personal characteristics.

Scenario 2: Causal — Although the wording is ambiguous, let's assume that we want to estimate the causal effect of winning versus losing an election on a candidate’s lifespan.

* Units:

The units differ. In the first scenario, we only care about (currently serving? currently alive even if retired? serving or will servce in the next few years?) US Senators. The second scenario is about political candidates (for all offices? outside the US?). We need to go back and forth with our boss/client/colleague. What exactly do they want?

Each row is a person, in both cases.

* Outcome:

The outcome variable of interest is the **age at death** (`death_age`) for each candidate. This is a continuous variable representing the candidate's total lifespan. Or is it `lived_after`? It is not obvious which is better. Alot depends on if `election_age` is included in the model. Probably best to try both and see how much your final answers differ.

* Covariates: 

Possible covariates include age at election, party affiliation, sex, vote margin, region, year of election, and other factors such as incumbent status or prior health, if available.
    
* Treatment: 

The treatment is whether or not the political candidate won or lost their election (`win_margin > 0`).


```{r}
tribble(
  ~`ID`, ~`Years Lived After Election`, ~`Election Result`, ~`Win Margin`, ~`Age at Election`, ~`Party`,
  "1",   "70",   "Win",   "2.5",   "55",   "Democrat",
  "2",   "56",   "Win",   "3.2",   "62",   "Republican",
  "...", "...", "...", "...", "...", "...",
  "10",  "65",   "Win",   "5.8",   "49",   "Democrat",
  "11",  "54",   "Win",   "1.7",   "67",   "Third",
  "...", "...", "...", "...", "...", "...",
  "N",   "56",   "Win",   "4.5",   "51",   "Republican"
) |>
  gt() |>
  tab_header(title = "Preceptor Table: Predicted Years Lived After Senate Election") |>
  tab_spanner(
    label = "Units",
    columns = c(`ID`)
  ) |>
  tab_spanner(
    label = "Outcome",
    columns = c(`Years Lived After Election`)
  ) |>
  tab_spanner(
    label = "Covariates",
    columns = c(`Win Margin`, `Age at Election`, `Party`)
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`ID`)) |>
  fmt_markdown(columns = everything())
```

Scenario 2 Preceptor Table:

```{r}
tribble(
  ~`ID`, ~`Years Lived After Election if Lose`, ~`Years Lived After Election if Win`, ~`Election Result`, ~`Age at Election`, ~`Win Margin`,
  "1",   "12*",   "18",    "Win",   "55",   "2.5",
  "2",   "24",    "26*",   "Lose",  "62",   "-3.2",
  "...", "...", "...", "...", "...", "...",
  "10",  "15*",   "21",    "Win",   "49",   "1.7",
  "11",  "11*",   "13",    "Win",   "67",   "0.5",
  "...", "...", "...", "...", "...", "...",
  "N",   "17",    "24*",   "Lose",  "51",   "-4.5"
) |>
  gt() |>
  tab_header(title = "Preceptor Table: Potential Years Lived After Election") |>
  tab_spanner(
    label = "Units",
    columns = c(`ID`)
  ) |>
  tab_spanner(
    label = "Potential Outcomes",
    columns = c(`Years Lived After Election if Lose`, `Years Lived After Election if Win`)
  ) |>
  tab_spanner(
    label = "Covariates",
    columns = c(`Age at Election`, `Win Margin`)
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`ID`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome which we do not observe."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  )
```

### Population Table 

Scenario 1 Population Table

```{r}
tribble(
  ~`Source`, ~`Year`, ~`ID`, ~`Years Lived After Election`, ~`Election Result`, ~`Win Margin`, ~`Age at Election`, ~`Party`, ~`Sex`,
  "...", "...", "...", "...", "...", "...", "...", "...", "...",
  "Data", "2002", "1", "77", "Win", "3.1", "51", "Republican", "Male",
  "Data", "1995", "2", "70", "Win", "2.7", "59", "Democrat", "Female",
  "...", "...", "...", "...", "...", "...", "...", "...", "...",
  "Preceptor Table", "2025", "10", "?", "Win", "2.5", "53", "Democrat", "Male",
  "Preceptor Table", "2025", "11", "?", "Win", "6.2", "48", "Republican", "Female",
  "...", "...", "...", "...", "...", "...", "...", "...", "..."
) |>
  gt() |>
  tab_header(title = "Population Table: Predicted Years Lived After Senate Election") |>
  tab_spanner(
    label = "Units/Time",
    columns = c(`Year`, `ID`)
  ) |>
  tab_spanner(
    label = "Outcome",
    columns = c(`Years Lived After Election`)
  ) |>
  tab_spanner(
    label = "Treatment",
    columns = c(`Election Result`)
  ) |>
  tab_spanner(
    label = "Covariates",
    columns = c(`Win Margin`, `Age at Election`, `Party`, `Sex`)
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Source`)) |>
  fmt_markdown(columns = everything())
```

Scenario 2 Population Table

```{r}
tribble(
  ~`Source`, ~`Year`, ~`ID`, ~`Years Lived After Election if Lose`, ~`Years Lived After Election if Win`, ~`Election Result`, ~`Win Margin`, ~`Age at Election`, ~`Party`, ~`Sex`,
  "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",
  "Data", "1992", "1", "14*", "19", "Win", "2.2", "60", "Republican", "Male",
  "Data", "2000", "2", "22", "26*", "Lose", "-3.8", "57", "Democrat", "Female",
  "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",
  "Preceptor Table", "2025", "10", "15*", "21", "Win", "1.5", "64", "Democrat", "Male",
  "Preceptor Table", "2025", "11", "17*", "16", "Lose", "-0.9", "70", "Republican", "Female",
  "...", "...", "...", "...", "...", "...", "...", "...", "...", "..."
) |>
  gt() |>
  tab_header(title = "Population Table: Potential Years Lived After Election") |>
  tab_spanner(
    label = "Unit/Time",
    columns = c(`Year`, `ID`)
  ) |>
  tab_spanner(
    label = "Potential Outcomes",
    columns = c(`Years Lived After Election if Lose`, `Years Lived After Election if Win`)
  ) |>
  tab_spanner(
    label = "Treatment",
    columns = c(`Election Result`)
  ) |>
  tab_spanner(
    label = "Covariates",
    columns = c(`Win Margin`, `Age at Election`, `Party`, `Sex`)
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Source`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome which we do not observe."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  )
```

Answer each question mentioned below:
  Describe the Population Table in your own words—what does each row represent in this context? Then, for each of the four key assumptions (validity, stability, representativeness, and unconfoundedness), provide both a definition in your own words and one specific counter-example for each assumption. 

Describe the Population Table in words:
  For scenario 1, each row corresponds to a candidate in a given year (e.g., a newly elected senator in 2025), showing their predicted years lived after election and covariates such as `age at election`, `party`, `sex`, and `win margin`.

  For scenario 2, each row corresponds to a candidate/year, with columns for potential years lived after winning or losing the election (even for outcomes not observed for every candidate), along with relevant covariates (`age`, `party`, `sex`, `win margin`, and treatment assignment).

Counter-example to the assumption of validity:
  In the governors dataset, if a candidate’s actual birth date or death date is unknown, researchers often substitute July 1 as the default day. This means some “years lived after election” values are based on estimates rather than exact dates. As a result, the outcome column is not defined consistently for every row, since some values represent true lifespans and others are just approximations. This weakens the validity assumption because not all entries have the same underlying meaning.

Counter-example to the assumption of stability:
  If the relationship between winning an election and longevity was different in earlier decades (e.g., in the 1950s) compared to today—perhaps because political office was more stressful or medical care for politicians was less advanced—then estimates using data from the past may not be stable over time.

Counter-example to the assumption of representativeness:
  If most of the data comes from governors in large states, but you want to make predictions or causal estimates for senators or candidates in smaller states, your results may not be representative. For example, candidates in less competitive or less affluent states may have different health backgrounds or access to healthcare.

Counter-example to the assumption of unconfoundedness:
  Imagine that candidates from wealthier backgrounds are more likely to win and also tend to live longer because they have better access to healthcare. If we don’t account for wealth in our model, it might look like winning the election leads to a longer life, when in fact it’s actually the candidate’s wealth that explains both winning and living longer.


### Modeling 

The outcome variable in this study is `lived_after`, which the years lived after the election. What kind of regression model should you use to predict or explain `lived_after`? Explain your choice based on what kind of variable `lived_after` is.

  You should use a linear regression model, because `death_age` is a continuous (numeric) variable. Linear regression is appropriate for predicting or explaining outcomes measured on a numeric scale like age.

