---
title: "Temperance and Governors"
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(gt)
library(primer.data)
library(tidymodels)
library(broom)
library(marginaleffects)

# Filter the data to make the assumptions of stability and unconfoundedness more reasonable.

x <- governors |>
  mutate(election_result = ifelse(win_margin > 0, "Win", "Lose")) |>
  filter(abs(win_margin) <= 5, year > 1945)

# Fit a model. This is our data generating mechanism.

fit_causal <- linear_reg() |>
  fit(lived_after ~ election_result + win_margin + election_age + party, data = x)
```

### Background

2) Imagine you are a researcher. You want to know if winning candidates live longer. 

```{r}
tribble(
  ~`Candidate`, ~`Years Lived (Lose)`, ~`Years Lived (Win)`, ~`Election Result`, ~`Age`, ~`Win Margin`, ~`Party`, ~`Sex`, ~`Other`,
  "Joe Smith",        "18*",   "23",   "Win",  "56", "7.2",  "Republican", "Male", "...",
  "David Jones",     "22",    "28*",  "Lose", "48", "-3.5",  "Democrat",   "Male", "...",
  "...",               "..." , "..." , "..." , "..." , "..." ,  "..." , "..." , "..."
) |>
  gt() |>
  tab_header(title = "Preceptor Table") |>
  tab_spanner(label = "Potential Outcomes", columns = c(`Years Lived (Lose)`, `Years Lived (Win)`)) |>
  tab_spanner(label = "Treatment", columns = c(`Election Result`)) |>
  tab_spanner(label = "Covariates", columns = c(`Age`, `Win Margin`, `Party`, `Sex`, `Other`)) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Candidate`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("Each row is a candidate running for election in 2025."),
    locations = cells_column_labels(columns = `Candidate`)
  ) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome that is not observed."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Years lived: Years lived after election, for both possible election results."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Other = additional covariates not listed here."),
    locations = cells_column_labels(columns = `Other`)
  )
```

```{r}
tribble(
  ~`Source`, ~`Candidate`, ~`Year`, ~`Years Lived (Lose)`, ~`Years Lived (Win)`, ~`Election Result`, ~`Win Margin`, ~`Age`, ~`Party`, ~`Sex`, ~`Other`,
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",
   "Data", "Earl Warren",  "1946", "18*", "23", "Win", "7.2", "56",  "Republican", "Male", "...",
   "Data", "George Wallace", "1946", "22", "28*", "Lose", "-3.5", "48", "Democrat", "Male", "...",
   "Data", "Nelson Rockefeller", "1946", "14*", "17", "Win", "9.0", "54",  "Republican", "Male", "...",
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", "...", 
   "Preceptor Table", "Joe Smith",  "2025",      "18*",   "23",   "Win",  "56", "7.2",  "Republican", "Male", "...",
     "Preceptor Table",  "David Jones",  "2025",    "22",    "28*",  "Lose", "48", "-3.5",  "Democrat",   "Male", "...",
   "...", "...", "...", "...", "...", "...", "...", "...", "...", "...",  "..."
) |>
  gt() |>
  tab_header(title = "Population Table") |>
  tab_spanner(label = "Unit/Time", columns = c(`Candidate`, `Year`)) |>
  tab_spanner(label = "Potential Outcomes", columns = c(`Years Lived (Lose)`, `Years Lived (Win)`)) |>
  tab_spanner(label = "Treatment", columns = c(`Election Result`)) |>
  tab_spanner(label = "Covariates", columns = c(`Win Margin`, `Age`,  `Party`, `Sex`, `Other`)) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(`Source`)) |>
  fmt_markdown(columns = everything()) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome that is not observed."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Years lived: Number of years lived after election for both possible outcomes."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  ) |>
  tab_footnote(
    footnote = md("Other = additional covariates not listed here. Age is at election."),
    locations = cells_column_spanners(spanners = "Covariates")
  )
```

## DGM

* Take a look at our fitted model. Use `tidy()` to examine the confidence intervals of the parameters. Are they any changes we might consider making?


```{r}
fit_causal
```

```{r}
fit_causal |> 
  tidy(conf.int = TRUE)
```

The model looks OK! You might think that the non-significant confidence interval for `partyThird` means that we should drop the `party` variable. After all, it (easily) includes zero. But that would be wrong! When you have a categorical variable like `party`, you should keep it as long as at least one of the 0/1 dummy variables is significant. In this case, `partyRepublican` is signicant --- Republicans seem to live longer than non-Republican --- so we should keep `party`.


* Use AI to create a nice looking table of the parameter values. (Only work on this question if your group is going very fast and has extra time.)

```{r}
fit_causal |> 
  tidy(conf.int = TRUE) |>
  gt() |>
  tab_header(
    title = "Linear Regression Results",
    subtitle = "Predicting Years Lived After Election"
  ) |>
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "SE",
    statistic = "t-statistic",
    p.value = "p-value",
    conf.low = "95% CI Lower",
    conf.high = "95% CI Upper"
  ) |>
  fmt_number(
    columns = c(estimate, std.error, statistic, conf.low, conf.high),
    decimals = 2
  ) |>
  fmt_scientific(
    columns = p.value,
    decimals = 2
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "lightgray"),
    locations = cells_body(
      columns = estimate
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = estimate,
      rows = p.value < 0.05
    )
  ) |>
  tab_footnote(
    footnote = "Bold estimates indicate statistical significance (p < 0.05)",
    locations = cells_column_labels(columns = estimate)
  ) |>
  tab_options(
    table.font.size = 12,
    heading.align = "center",
    column_labels.border.top.style = "solid",
    column_labels.border.bottom.style = "solid",
    table_body.border.bottom.style = "solid"
  )
``` 

* Use AI to create LaTeX code for our DGM. (Only work on this question if your group is going very fast and has extra time.)

$$
\begin{align}
\text{lived\_after} &= 67.56 + 8.63 \cdot \text{election\_result}_{\text{Win}} - 1.46 \cdot \text{win\_margin} \nonumber \\
&\quad - 0.89 \cdot \text{election\_age} + 4.04 \cdot \text{party}_{\text{Republican}} - 9.50 \cdot \text{party}_{\text{Third party}}
\end{align}
$$

## Questions and Answers

Recall our original question: Do winning candidates live longer? Well, do they? And do other candidate characteristics matter?

* First, let's look at individual variables which are included in our DGM. Hint: Use `plot_predictions()`

```{r}
fit_causal |> 
  predictions()
```

```{r}
fit_causal |> 
  plot_predictions(condition = c("election_result"))
```

```{r}
fit_causal |> 
  plot_predictions(condition = c("party"))
```

```{r}
fit_causal |> 
  plot_predictions(condition = c("election_age"))
```


* Second, combine several variables --- and think about which ones you should select --- into a nice looking plot. There is no wrong answer! But some plots look better than others.

```{r}
fit_causal |> 
  plot_predictions(condition = c("election_age",  "election_result", "party"))
```

* If there is time, use the `draw = FALSE` option to `plot_predictions()` to pull out the raw data. Then, tell an AI about that tibble and ask for some nice ggplot code. Pay special attention to the subtitle.



## Humility

* What is, truly, your best guess as to the causal effect of winning an election on longevity?

My answer: I predict that candidate who win an election live a few years longer than they would have otherwise, but I would not be surprised if, in fact, the effect is actually negative. My best guess is 3 years +/-5.

Why is my 



