## Day 2: Courage

```{r}
#| message: false

library(tidyverse)
library(tidymodels)
library(broom)
library(marginaleffects)
library(easystats)
library(gt)

x <- read_csv("https://raw.githubusercontent.com/PPBDS/bootcamp/refs/heads/master/_internal/data/STAR.csv",
              show_col_types = FALSE) |>
                mutate(kind = recode(classtype,
                       `1` = "small",
                       `2` = "regular",
                       `3` = "regular_with_aid"))
```

<!-- DK: Change code and other stuff to be consistent.  -->


## Background Information

Frederick Mosteller (1995) “The Tennessee study of class size in the early school grades.” The Future of Children, vol. 5, no. 2, pp. 113–127. Data cleaned and discussed in "Quantitative Social Science: An Introduction" by Kosuke Imai.

"The Tennessee class size project is a three-phase study designed to determine the effect of smaller class size in the earliest grades on short-term and long-term pupil performance. The first phase of this project, termed Project STAR (for Student-Teacher Achievement Ratio), was begun in 1985, when Lamar Alexander was governor of Tennessee. Governor Alexander, who later served as secretary of education in the cabinet of President George Bush, had made education a top priority for his second term.

The legislature and the educational community of Tennessee were mindful of a promising study of the benefits of small class size carried out in nearby Indiana, but were also aware of the costs associated with additional classrooms and teachers. Wishing to obtain data on the effectiveness of reduced class size before committing additional funds, the Tennessee legislature authorized this four-year study in which results obtained in kindergarten, first, second, and third grade classrooms of 13 to 17 pupils were compared with those obtained in classrooms of 22 to 25 pupils and in classrooms of this larger size where the teacher was assisted by a paid aide. Both standardized and curriculum-based tests were used to assess and compare the performance of some 6,500 pupils in about 330 classrooms at approximately 80 schools in the areas of reading, mathematics, and basic study skills. After four years, it was clear that smaller classes did produce substantial improvement in early learning and cognitive studies and that the effect of small class size on the achievement of minority children was initially about double that observed for majority children, but in later years, it was about the same.

The second phase of the project, called the Lasting Benefits Study, was begun in 1989 to determine whether these perceived benefits persisted. Observations made as a part of this phase confirmed that the children who were originally enrolled in smaller classes continued to perform better than their grade-mates (whose school experience had begun in larger classes) when they were returned to regular-sized classes in later grades. Under the third phase, Project Challenge, the 17 economically poorest school districts were given small classes in kindergarten, first, second, and third grades. These districts improved their end-of-year standing in rank among the 139 districts from well below average to above average in reading and mathematics. This article briefly summarizes the Tennessee class size project, a controlled experiment which is one of the most important educational investigations ever carried out and illustrates the kind and magnitude of research needed in the field of education to strengthen schools."

### Scenarios

1) Imagine that you are an elementary school principal in Nashville in 2000. You want to predict whether reducing the students per class will likely boost math achievement next year in your school.

2) Imagine you work for the Texas Department of Education. You want to use the STAR data to predict what might happen if Dallas adopted small class sizes, but you are worried about whether the findings from 1980s Tennessee would apply to Dallas today.

<!-- AT: Check examples. -->

### Counter-Arguments

## Scenario 1

* Counter-argument to the assumption of validity:
>**The definition of “small class” or the math achievement test used in Nashville in 2000 may not perfectly match those from the STAR study in Tennessee, so the measured outcome may not be directly comparable.**

* Counter-argument to the assumption of stability:
>**The effect of reducing class size on math achievement could be different in Nashville in 2000 than it was in Tennessee in the 1980s, due to differences in students, teachers, or school environments.**

* Counter-argument to the assumption of representativeness:
>**Nashville’s students in 2000 may differ in key ways from students in the STAR study—such as demographics, family backgrounds, or school funding—which could limit how well the STAR results apply to Nashville.**

* Counter-argument to the assumption of unconfoundedness:
>**In practice, it may not be possible to randomly assign students to small classes in Nashville, and other factors (like which schools or students are chosen for small classes) could affect math scores, making it hard to isolate the causal effect.**

## Scenario 2
* Counter-argument to the assumption of validity:
>**“Small class” and “math achievement” in Dallas today might be defined or measured differently than in 1980s Tennessee, so using STAR results directly may not be valid.**

* Counter-argument to the assumption of stability:
>**The relationship between class size and math achievement may have changed over time or may differ across locations, so the effect observed in 1980s Tennessee may not be stable in Dallas today.**

* Counter-argument to the assumption of representativeness:
>**The STAR study participants from Tennessee might not be representative of Dallas students today—differences in socioeconomic status, educational resources, or community factors could affect generalizability.**

* Counter-argument to the assumption of unconfoundedness:
>**If Dallas does not implement small classes randomly, schools or students selected for the intervention might differ in important ways, making it difficult to separate the effect of class size from other influences.**


### Example Preceptor Tables
  
```{r}
tibble(
  ID = c("1", "2", "...", "10", "11", "...", "N"),
  score_regular = c("706*", "565", "...", "654*", "547*", "...", "568"),
  score_small = c("638", "620*", "...", "710", "567", "...", "743*"),
  observed_class = c("Small", "Regular", "...", "Small", "Small", "...", "Regular"),
  sex = c("female", "male", "...", "male", "female", "...", "female"),
  race = c("white", "black", "...", "black", "white", "...", "white"),
  school = c("A", "B", "...", "C", "A", "...", "B")
) |>
  gt() |>
  tab_header(title = "Preceptor Table: Potential Outcomes by Class Size") |>
  cols_label(
    ID = md("ID"),
    score_regular = md("Score if Regular Class"),
    score_small = md("Score if Small Class"),
    observed_class = md("Observed Class Size"),
    sex = md("Sex"),
    race = md("Race"),
    school = md("School")
  ) |>
  tab_spanner(label = "Potential Outcomes", columns = c(score_regular, score_small)) |>
  tab_spanner(label = "Covariates", columns = c(sex, race, school)) |>
  fmt_markdown(columns = everything()) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(ID)) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome we do not observe."),
    locations = cells_column_spanners(spanners = "Potential Outcomes")
  )
```


```{r}
tibble(
  ID = c("1", "2", "...", "10", "11", "...", "N"),
  score_regular = c("685", "672*", "...", "689", "661", "...", "670*"),
  score_small   = c("700*", "679", "...", "695*", "675*", "...", "678"),
  gain_if_small = c("15*", "7", "...", "6*", "14*", "...", "8"),
  observed_class = c("Regular", "Small", "...", "Regular", "Regular", "...", "Small"),
  race = c("white", "black", "...", "white", "black", "...", "white"),
  yearssmall = c("0", "4", "...", "0", "0", "...", "4"),
  school = c("A", "B", "...", "C", "A", "...", "B")) |>
  gt() |>
  tab_header(title = "Preceptor Table: Causal Effect of Small Class Size") |>
  cols_label(
    ID = md("ID"),
    score_regular = md("Score if Regular Class"),
    score_small = md("Score if Small Class"),
    gain_if_small = md("Gain if Small Class"),
    observed_class = md("Observed Class Size"),
    race = md("Race"),
    yearssmall = md("Years in Small Class"),
    school = md("School")
  ) |>
  tab_spanner(label = "Potential Outcomes", columns = c(score_regular, score_small)) |>
  tab_spanner(label = "Estimated Causal Effect", columns = c(gain_if_small)) |>
  tab_spanner(label = "Covariates", columns = c(observed_class, race, yearssmall, school)) |>
  fmt_markdown(columns = everything()) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(ID)) |>
  tab_footnote(
    footnote = md("A * indicates a potential outcome or gain we do not observe."),
    locations = cells_column_spanners(spanners = c("Potential Outcomes", "Estimated Causal Effect"))
  )

```

## Example Population Tables

```{r}
tibble(
  ID = c("1", "2", "...", "10", "11", "...", "N"),
  math_score = c("638", "565", "...", "654", "547", "...", "743"),
  class_size = c("Small", "Regular", "...", "Small", "Small", "...", "Regular"),
  sex = c("female", "male", "...", "male", "female", "...", "female"),
  race = c("white", "black", "...", "black", "white", "...", "white"),
  yearssmall = c("4", "0", "...", "4", "4", "...", "0"),
  school = c("A", "B", "...", "C", "A", "...", "B")) |>
  gt() |>
  tab_header(title = "Population Table: Observed Outcomes and Covariates") |>
  cols_label(
    ID = md("ID"),
    math_score = md("Observed Math Score"),
    class_size = md("Observed Class Size"),
    sex = md("Sex"),
    race = md("Race"),
    yearssmall = md("Years in Small Class"),
    school = md("School")
  ) |>
  fmt_markdown(columns = everything()) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(ID)) |>
  tab_spanner(label = "Observed Data", columns = c(math_score, class_size, sex, race, yearssmall, school))
```


```{r}
tibble(
  ID = c("1", "2", "...", "10", "11", "...", "N"),
  math_score = c("685", "679", "...", "689", "661", "...", "678"),
  class_size = c("Regular", "Small", "...", "Regular", "Regular", "...", "Small"),
  race = c("white", "black", "...", "white", "black", "...", "white"),
  yearssmall = c("0", "4", "...", "0", "0", "...", "4")) |>
  gt() |>
  tab_header(title = "Population Table: Alternate Version") |>
  cols_label(
    ID = md("ID"),
    math_score = md("Observed Math Score"),
    class_size = md("Observed Class Size"),
    race = md("Race"),
    yearssmall = md("Years in Small Class")
  ) |>
  fmt_markdown(columns = everything()) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(ID)) |>
  tab_spanner(label = "Observed Data", columns = c(math_score, class_size, race, yearssmall))
```


